---
title: "Predicting Urban NO\\textsubscript{2} Concentrations from Low-Cost Sensor Arrays"
subtitle: "STAT 24620 / FINM 34700 / STAT 32950 --- Group Data Project, Spring 2026"
author: "Cole Ginter, Vidhi Jain, Sunil Trivedi, Maisha Miah"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: false
    number_sections: false
    fig_caption: true
    keep_tex: false
geometry: margin=0.85in
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \floatplacement{figure}{H}
  - \setlength{\parskip}{2pt}
  - \setlength{\textfloatsep}{6pt plus 1pt minus 2pt}
  - \setlength{\intextsep}{6pt plus 1pt minus 2pt}
---

```{r setup, include=FALSE}
# Global knitr options: hide code throughout the body; show results, figures, tables.
# All code is reproduced in the Code Appendix at the end.
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 6.5,
  fig.height = 3,
  dpi       = 120,
  out.width = "92%"
)

# Packages
library(tidyverse)   # dplyr, ggplot2, tidyr, readr
library(lubridate)   # datetime arithmetic
library(patchwork)   # combining ggplots
library(glmnet)      # ridge + lasso
library(e1071)       # skewness()
library(zoo)         # rolling means
library(knitr)       # kable()
library(kableExtra)  # kable_styling()

set.seed(42)
```

**AI acknowledgment:** Claude Code (Anthropic) was used for coding assistance and debugging. All analysis design, interpretation, and written explanations are the authors' own work.

## Section 1: Research Question

For our final project we seek to answer the question: Can a five-sensor electrochemical array (PT08_S1--S5) accurately predict hourly NO\textsubscript{2} concentrations at an Italian urban monitoring site, and does dimensionality reduction via PCA improve prediction relative to directly regressing on the raw sensor readings? Air-quality monitoring increasingly relies on low-cost sensors alongside certified reference analyzers. This is an adaptation of the research question originally proposed in De Vito et al., 2008 which allows us to analyze a real-world statistics problem while using methods taught in this course. Firstly, if a small number of principal components captures most of the sensor variance, a PC regression may be more stable than a full sensor regression, (especially since we hypothesize that some sensors are collinear and subject to environmental drift). To answer this question we will compare OLS, Ridge, Lasso, and PC regression, all augmented with meteorological covariates, on a chronological hold-out test set. Along these lines, the variable groups that motivate this question are:

-   Predictors: Sensors (`PT08_S1_CO`, `PT08_S2_NMHC`, `PT08_S3_NOx`, `PT08_S4_NO2`, `PT08_S5_O3`)
-   Covaraites: Weather (`T` (°C), `RH` (%), `AH` (g/m\textsuperscript{3}))
-   Reference (to contextualize response): Measurements (`CO_GT`, `C6H6_GT`, `NOx_GT`, `NO2_GT`)
-   Response: `NO2_GT` (µg/m\textsuperscript{3}).

## Section 2: Discriptive Data Exploration and Visualization

```{r load-data, results='hide'}
setwd("/Users/gregoryginter/Desktop/UChicago MSFM/FINM 34700/Final Project")
df <- read.csv("air_quality_italy.csv",
               stringsAsFactors = FALSE)

# Diagnostic checks (suppressed in body; visible in Code Appendix)
cat("Shape:", nrow(df), "rows x", ncol(df), "cols\n")
print(sapply(df, class))
print(colSums(is.na(df)))

num_cols   <- names(df)[sapply(df, is.numeric)]
neg200_hit <- sapply(df[num_cols], function(x) any(x == -200, na.rm = TRUE))
neg200_in  <- names(neg200_hit)[neg200_hit]
if (length(neg200_in)) print(neg200_in) else cat("None -- clean.\n")

# Build proper datetime index
df$timestamp <- as.POSIXct(df$Date, tz = "UTC") + lubridate::hours(df$Hour)
df <- df[order(df$timestamp), ]
rownames(df) <- NULL

cat("\nDate range:", format(min(df$timestamp)), "to", format(max(df$timestamp)), "\n")
cat("Total hours:", nrow(df), "\n")

SENSORS <- c("PT08_S1_CO", "PT08_S2_NMHC", "PT08_S3_NOx", "PT08_S4_NO2", "PT08_S5_O3")
GT_VARS <- c("CO_GT", "C6H6_GT", "NOx_GT", "NO2_GT")
WEATHER <- c("T", "RH", "AH")
TARGET  <- "NO2_GT"
```

```{r fig1-eda, fig.width=7, fig.height=2, fig.cap="NO$_2$ reference concentration over time (left) and distribution on raw vs.\\ log scale (right)."}
ALL_VARS <- c(SENSORS, GT_VARS, WEATHER)

df$NO2_roll <- zoo::rollmean(df$NO2_GT, k = 168, fill = NA, align = "right")

p_ts <- ggplot(df, aes(x = timestamp)) +
  geom_line(aes(y = NO2_GT, color = "Hourly"), linewidth = 0.2, alpha = 0.55) +
  geom_line(aes(y = NO2_roll, color = "7-day rolling mean"), linewidth = 0.7) +
  scale_color_manual(name = NULL,
                     values = c("Hourly" = "steelblue",
                                "7-day rolling mean" = "darkblue")) +
  labs(x = "Date", y = expression(NO[2]~GT~(mu*g/m^3)),
       title = expression(NO[2]~"Over Time")) +
  theme_minimal(base_size = 8) +
  theme(legend.position = "bottom",
        legend.margin = margin(0,0,0,0),
        legend.box.spacing = unit(2, "pt"),
        axis.text.x = element_text(angle = 30, hjust = 1),
        plot.title = element_text(size = 9))

log_vals <- log(df$NO2_GT[df$NO2_GT > 0 & !is.na(df$NO2_GT)])

p_dist <- ggplot() +
  geom_histogram(aes(x = df$NO2_GT, y = after_stat(density), fill = "Raw NO2_GT"),
                 bins = 60, alpha = 0.6) +
  geom_histogram(aes(x = log_vals * 30, y = after_stat(density) / 30,
                     fill = "log(NO2_GT) (rescaled)"),
                 bins = 60, alpha = 0.45) +
  scale_fill_manual(name = NULL,
                    values = c("Raw NO2_GT" = "steelblue",
                               "log(NO2_GT) (rescaled)" = "darkorange")) +
  labs(x = expression(NO[2]~GT~(mu*g/m^3)), y = "Density",
       title = expression("Distribution of"~NO[2]~GT)) +
  theme_minimal(base_size = 8) +
  theme(legend.position = "bottom",
        legend.margin = margin(0,0,0,0),
        legend.box.spacing = unit(2, "pt"),
        plot.title = element_text(size = 9))

p_ts + p_dist
```

The time-series panel shows peaks in the morning and evening likely attributable to work traffic as well as a seasonal drift upwards in the fall and winter periods when boundary-layer mixing is suppressed and a drift downwards in the spring and summer period. The density plot shows that the raw `NO2_GT` is mildly right-skewed (`r sprintf("%.2f", e1071::skewness(df$NO2_GT, na.rm = TRUE))`) which lead us to try a log transformation but this over-corrects to a left-skewed distribution (`r sprintf("%.2f", e1071::skewness(log_vals, na.rm = TRUE))`). Because the log transformation does not improve symmetry, we fit all supervised models on the raw concentration scale and further discussed this in limitation section.


## Section 3: Analysis of Data Structure

We begin by providing the summary statsitics of the data used in the model to gain a better understanding of the overall shape of the data used.

```{r summary-stats}
sum_tab <- do.call(rbind, lapply(ALL_VARS, function(v) {
  x <- df[[v]]
  x <- x[!is.na(x)]
  data.frame(
    Variable = v,
    N        = length(x),
    Mean     = mean(x),
    SD       = sd(x),
    Min      = min(x),
    Q1       = quantile(x, 0.25),
    Median   = median(x),
    Q3       = quantile(x, 0.75),
    Max      = max(x),
    Skewness = e1071::skewness(x),
    check.names = FALSE,
    row.names   = NULL
  )
}))

sum_tab[, 3:10] <- round(sum_tab[, 3:10], 2)

kable(sum_tab,
      caption  = "Summary Statistics of Model Variables",
      booktabs = TRUE,
      col.names = c("Variable", "N", "Mean", "SD", "Min", "Q1", "Median", "Q3", "Max", "Skewness")) |>
  kable_styling(font_size = 7, latex_options = "hold_position")
```

The sensors seem to have roughly the same scale tending to lie between  650 and 1,700 and they also have a moderate right skew, which could be attributed to occasional traffic-spike outliers. The reference variables use different  units: `CO_GT` is on a 0--10 mg/m\textsuperscript{3} scale while `NOx_GT` and `NO2_GT` reach into the hundreds in units of µg/m\textsuperscript{3}. Temperature (`T`) averages `r round(mean(df$T, na.rm=TRUE), 1)` °C with a wide range reflecting the full yearly seasonal cycle. Relative humidity (`RH`) expectedly hovers relatively tightly and uniformly around 50% on average. The skewness values confirm the mild right skew of `NO2_GT` discussed in the time-series panel above. Next we analyze the correlation of the variables. 

```{r fig2-corr, fig.width=6.5, fig.height=4.3, fig.cap="Pearson correlation heatmap across sensors (top-left block), reference analyzers (middle block), and weather (bottom-right block). Block separators are drawn in black."}
corr_cols <- c(SENSORS, GT_VARS, WEATHER)
corr_mat  <- cor(df[, corr_cols], use = "pairwise.complete.obs")

corr_df <- as.data.frame(as.table(corr_mat))
names(corr_df) <- c("Var1", "Var2", "value")
corr_df$Var1 <- factor(corr_df$Var1, levels = corr_cols)
corr_df$Var2 <- factor(corr_df$Var2, levels = rev(corr_cols))

block_breaks <- c(length(SENSORS), length(SENSORS) + length(GT_VARS))

ggplot(corr_df, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white", linewidth = 0.2) +
  geom_text(aes(label = sprintf("%.2f", value)), size = 2.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  geom_vline(xintercept = block_breaks + 0.5, linewidth = 0.5) +
  geom_hline(yintercept = length(corr_cols) - block_breaks + 0.5, linewidth = 0.5) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.key.width = unit(8, "pt"))
```

This correlation matrix supports the use of PCA as the sensor section of the matrix has high correlation with each other which implies the matrix has low intrinsic rank, which PCA could help alleviate. The matrix also supports our approach to the research question as there is cross-block correlation between the sensors and `NO2_GT`, supporting sensor-based prediction.

Because we are using hourly time-series data that has strong autocorrelation and seasonal affects, a random split would leak future information into the past. Therefore, we use the first 75% of the data for the training set and the remaining 25% form the test set. All the subsequent preprocessing statistics (scaling means/SDs, PCA directions, regularization hyperparameters) are estimated using the training data only and only applied to test data without re-fitting.

```{r split}
n         <- nrow(df)
split_idx <- floor(0.75 * n)
train <- df[1:split_idx, ]
test  <- df[(split_idx + 1):n, ]

split_tab <- data.frame(
  Set   = c("Training", "Test"),
  Rows  = c(nrow(train), nrow(test)),
  Start = c(format(min(train$timestamp), "%Y-%m-%d"),
            format(min(test$timestamp),  "%Y-%m-%d")),
  End   = c(format(max(train$timestamp), "%Y-%m-%d"),
            format(max(test$timestamp),  "%Y-%m-%d"))
)

kable(split_tab,
      caption  = "Chronological 75/25 train/test split.",
      booktabs = TRUE,
      col.names = c("Set", "Rows", "Start", "End")) |>
  kable_styling(latex_options = "hold_position")
```

## Section 4: PCA on the Sensor Block

We run PCA exclusively on the five PT08 sensor variables. Weather variables are excluded so that the structural story stays clean: the question here is *how many independent axes of variation exist in the sensor array itself?* Weather enters only as covariates in the supervised stage. Standardization (zero mean, unit variance) uses training-set statistics only and is applied identically to the test sensors.

```{r pca-fit, results='hide'}
train_sensors <- as.matrix(train[, SENSORS])
test_sensors  <- as.matrix(test[, SENSORS])

sens_mean <- colMeans(train_sensors)
sens_sd   <- apply(train_sensors, 2, sd)

train_sensors_sc <- scale(train_sensors, center = sens_mean, scale = sens_sd)
test_sensors_sc  <- scale(test_sensors,  center = sens_mean, scale = sens_sd)

pca <- prcomp(train_sensors_sc, center = FALSE, scale. = FALSE)
train_pcs <- pca$x
test_pcs  <- as.matrix(test_sensors_sc) %*% pca$rotation

var_exp <- pca$sdev^2 / sum(pca$sdev^2)
cum_var <- cumsum(var_exp)

loadings <- pca$rotation
colnames(loadings) <- paste0("PC", 1:ncol(loadings))

# Suppressed; figure and discussion convey the same information
cat("Variance explained per PC:\n")
for (i in seq_along(var_exp)) {
  cat(sprintf("  PC%d: %.3f  (cumulative: %.3f)\n", i, var_exp[i], cum_var[i]))
}
cat("\nLoadings matrix:\n")
print(round(loadings, 3))
```

```{r fig3-scree, fig.width=7, fig.height=2.6, fig.cap="Scree plot (left) and PC1/PC2 sensor loadings (right)."}
scree_df <- data.frame(PC = factor(paste0("PC", 1:5), levels = paste0("PC", 1:5)),
                       var = var_exp, cum = cum_var)

p_scree <- ggplot(scree_df, aes(x = PC)) +
  geom_col(aes(y = var), fill = "steelblue", alpha = 0.75) +
  geom_line(aes(y = cum, group = 1), color = "darkred", linewidth = 0.7) +
  geom_point(aes(y = cum), color = "darkred", size = 1.5) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "gray40") +
  annotate("text", x = 4.7, y = 0.92, label = "90%", color = "gray40", size = 2.5) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(y = "Variance explained", title = "Scree", x = NULL) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))

sensor_labels <- c(
  "PT08_S1_CO"   = "S1\n(CO)",
  "PT08_S2_NMHC" = "S2\n(NMHC)",
  "PT08_S3_NOx"  = "S3\n(NOx)",
  "PT08_S4_NO2"  = "S4\n(NO2)",
  "PT08_S5_O3"   = "S5\n(O3)"
)

load_df <- as.data.frame(loadings[, 1:2])
load_df$sensor <- factor(rownames(load_df), levels = SENSORS)
load_long <- tidyr::pivot_longer(load_df, cols = c("PC1", "PC2"),
                                 names_to = "component", values_to = "loading")

p_load <- ggplot(load_long,
                 aes(x = sensor, y = loading, fill = component)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "darkorange"),
                    name = NULL) +
  scale_x_discrete(labels = sensor_labels) +
  labs(y = "Loading", x = NULL, title = "PC1 / PC2 loadings") +
  theme_minimal(base_size = 8) +
  theme(legend.position = "bottom",
        legend.margin = margin(0,0,0,0),
        legend.box.spacing = unit(2, "pt"),
        plot.title = element_text(size = 9))

p_scree + p_load
```

**PC1 (`r sprintf("%.0f%%", 100*var_exp[1])` of variance)** loads negatively and nearly equally on four of the five sensors ($\approx +0.42$ to $+0.47$), but *positively* on `PT08_S3_NOx` ($-0.43$). This sign flip is physically meaningful: PT08_S3 is a metal-oxide sensor tuned to NOx, which responds inversely to ambient NO\textsubscript{2} oxidising conditions compared to the other sensors. PC1 therefore represents the overall combustion/traffic pollution load, with PT08_S3 providing a counteracting signal. **PC2 (`r sprintf("%.0f%%", 100*var_exp[2])` of variance)** is strongly dominated by `PT08_S4_NO2` ($-0.84$), with `PT08_S3_NOx` ($-0.39$) and `PT08_S5_O3` ($+0.38$) as secondary contributors --- reflecting the chemical antagonism between NO\textsubscript{2} and O\textsubscript{3} in urban photochemistry. Together PC1 and PC2 explain `r sprintf("%.0f%%", 100*cum_var[2])` of sensor variance; PC3--PC5 each capture less than 6%. The low-rank structure motivates PC regression: projecting onto the first few PCs substitutes correlated raw sensors with orthogonal summary axes, which can stabilise regression coefficients. Whether this leads to better out-of-sample prediction depends on whether the discarded PCs carry useful signal --- tested in Section 5.

## Section 5: Supervised Models

We fit four models predicting `NO2_GT` (raw scale), each using the five sensors plus `T`, `RH`, `AH` as predictors. All predictors are standardised using a separate scaler fit on training data only. **OLS** (baseline, no regularization); **Ridge** ($L_2$ penalty, $\lambda$ by 5-fold CV); **Lasso** ($L_1$ penalty, $\lambda$ by 5-fold CV); **PC regression** (OLS on first $k$ PCs + weather, $k$ by 5-fold CV). This R implementation uses `glmnet`, whose penalty $\lambda$ is parameterised differently from `scikit-learn`'s $\alpha$ (glmnet scales the loss by $1/n$); optimal numerical values therefore differ from a Python implementation, but the methodology is identical.

```{r build-features, results='hide'}
FEATURES <- c(SENSORS, WEATHER)

X_train_raw <- as.matrix(train[, FEATURES])
X_test_raw  <- as.matrix(test[, FEATURES])

sup_mean <- colMeans(X_train_raw)
sup_sd   <- apply(X_train_raw, 2, sd)

X_train <- scale(X_train_raw, center = sup_mean, scale = sup_sd)
X_test  <- scale(X_test_raw,  center = sup_mean, scale = sup_sd)

y_train <- train[[TARGET]]
y_test  <- test[[TARGET]]
```

```{r fit-ols-ridge-lasso, results='hide'}
# --- OLS ---
ols_df_train <- data.frame(y = y_train, X_train)
ols_fit      <- lm(y ~ ., data = ols_df_train)
ols_train_pred <- predict(ols_fit, newdata = ols_df_train)
ols_test_pred  <- predict(ols_fit, newdata = data.frame(X_test))

# --- Ridge (5-fold CV on train) ---
ridge_cv <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 5,
                      type.measure = "mse", standardize = FALSE)
ridge_lambda     <- ridge_cv$lambda.min
ridge_train_pred <- as.numeric(predict(ridge_cv, newx = X_train, s = "lambda.min"))
ridge_test_pred  <- as.numeric(predict(ridge_cv, newx = X_test,  s = "lambda.min"))

# --- Lasso (5-fold CV on train) ---
lasso_cv <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 5,
                      type.measure = "mse", standardize = FALSE)
lasso_lambda     <- lasso_cv$lambda.min
lasso_train_pred <- as.numeric(predict(lasso_cv, newx = X_train, s = "lambda.min"))
lasso_test_pred  <- as.numeric(predict(lasso_cv, newx = X_test,  s = "lambda.min"))

lasso_coefs_full <- as.numeric(coef(lasso_cv, s = "lambda.min"))
names(lasso_coefs_full) <- c("(Intercept)", FEATURES)
lasso_coefs <- lasso_coefs_full[-1]

zeroed   <- names(lasso_coefs)[lasso_coefs == 0]
retained <- names(lasso_coefs)[lasso_coefs != 0]
```

```{r fit-pcr, results='hide'}
# --- PC Regression: choose k by 5-fold CV on train (preserving time order) ---
train_weather_raw <- as.matrix(train[, WEATHER])
test_weather_raw  <- as.matrix(test[, WEATHER])
w_mean <- colMeans(train_weather_raw)
w_sd   <- apply(train_weather_raw, 2, sd)
train_weather_sc <- scale(train_weather_raw, center = w_mean, scale = w_sd)
test_weather_sc  <- scale(test_weather_raw,  center = w_mean, scale = w_sd)

n_train  <- nrow(X_train)
fold_id  <- ceiling(seq_len(n_train) / (n_train / 5))
fold_id  <- pmin(fold_id, 5L)

cv_rmse <- numeric(5)
for (k in 1:5) {
  fold_rmse <- numeric(5)
  X_pc_full <- cbind(train_pcs[, 1:k, drop = FALSE], train_weather_sc)
  for (f in 1:5) {
    val_idx  <- which(fold_id == f)
    tr_idx   <- setdiff(seq_len(n_train), val_idx)
    fit      <- lm.fit(cbind(1, X_pc_full[tr_idx, , drop = FALSE]), y_train[tr_idx])
    pred     <- cbind(1, X_pc_full[val_idx, , drop = FALSE]) %*% fit$coefficients
    fold_rmse[f] <- sqrt(mean((y_train[val_idx] - pred)^2))
  }
  cv_rmse[k] <- mean(fold_rmse)
}

best_k <- which.min(cv_rmse)

X_pc_train_best <- cbind(train_pcs[, 1:best_k, drop = FALSE], train_weather_sc)
X_pc_test_best  <- cbind(test_pcs[,  1:best_k, drop = FALSE], test_weather_sc)
colnames(X_pc_train_best) <- c(paste0("PC", 1:best_k), WEATHER)
colnames(X_pc_test_best)  <- colnames(X_pc_train_best)

pcr_df_train <- data.frame(y = y_train, X_pc_train_best)
pcr_fit      <- lm(y ~ ., data = pcr_df_train)
pcr_train_pred <- predict(pcr_fit, newdata = pcr_df_train)
pcr_test_pred  <- predict(pcr_fit, newdata = data.frame(X_pc_test_best))
```

## Section 6: Comparison

```{r table1-results}
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

results <- data.frame(
  Model = c(
    "OLS (baseline)",
    sprintf("Ridge (lambda=%.3f)",  ridge_lambda),
    sprintf("Lasso (lambda=%.4f)",  lasso_lambda),
    sprintf("PC Regression (k=%d)", best_k)
  ),
  `Train RMSE` = c(rmse(y_train, ols_train_pred),
                   rmse(y_train, ridge_train_pred),
                   rmse(y_train, lasso_train_pred),
                   rmse(y_train, pcr_train_pred)),
  `Test RMSE`  = c(rmse(y_test, ols_test_pred),
                   rmse(y_test, ridge_test_pred),
                   rmse(y_test, lasso_test_pred),
                   rmse(y_test, pcr_test_pred)),
  `Test R2`    = c(r2(y_test, ols_test_pred),
                   r2(y_test, ridge_test_pred),
                   r2(y_test, lasso_test_pred),
                   r2(y_test, pcr_test_pred)),
  check.names = FALSE
)
results[, 2:4] <- round(results[, 2:4], 3)

kable(results,
      caption  = "Model comparison on the chronological hold-out test set.",
      booktabs = TRUE,
      col.names = c("Model", "Train RMSE", "Test RMSE", "Test $R^2$"),
      escape    = FALSE)
```

```{r fig4-pred-vs-actual, fig.width=6.5, fig.height=5, fig.cap="Predicted vs.\\ actual NO$_2$ on the chronological test set, by model."}
model_preds <- list(
  list(name = "OLS (baseline)",                              pred = ols_test_pred,   color = "steelblue"),
  list(name = sprintf("Ridge (lambda=%.3f)", ridge_lambda),  pred = ridge_test_pred, color = "forestgreen"),
  list(name = sprintf("Lasso (lambda=%.4f)", lasso_lambda),  pred = lasso_test_pred, color = "darkorange"),
  list(name = sprintf("PC Regression (k=%d)", best_k),       pred = pcr_test_pred,   color = "#DC143C")
)

plot_df <- do.call(rbind, lapply(model_preds, function(m) {
  data.frame(model     = m$name,
             color     = m$color,
             actual    = y_test,
             predicted = m$pred,
             stringsAsFactors = FALSE)
}))
plot_df$model <- factor(plot_df$model, levels = sapply(model_preds, function(m) m$name))

ann_df <- plot_df %>%
  group_by(model) %>%
  summarise(rmse_val = sqrt(mean((actual - predicted)^2)),
            r2_val   = 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2),
            .groups  = "drop") %>%
  mutate(label = sprintf("RMSE = %.2f, R^2 = %.3f", rmse_val, r2_val))

lo <- min(c(plot_df$actual, plot_df$predicted)) - 5
hi <- max(c(plot_df$actual, plot_df$predicted)) + 5

ggplot(plot_df, aes(x = actual, y = predicted)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "black", linewidth = 0.3) +
  geom_point(aes(color = model), alpha = 0.18, size = 0.4, show.legend = FALSE) +
  scale_color_manual(values = setNames(sapply(model_preds, function(m) m$color),
                                       sapply(model_preds, function(m) m$name))) +
  facet_wrap(~ model, ncol = 2) +
  geom_text(data = ann_df,
            aes(x = lo + 5, y = hi - 10, label = label),
            hjust = 0, vjust = 1, size = 2.4, inherit.aes = FALSE) +
  coord_fixed(xlim = c(lo, hi), ylim = c(lo, hi)) +
  labs(x = expression("Actual"~NO[2]~GT),
       y = expression("Predicted"~NO[2]~GT)) +
  theme_minimal(base_size = 8) +
  theme(strip.text = element_text(face = "bold", size = 8),
        panel.spacing = unit(4, "pt"))
```

**Discussion.** All four models achieve nearly identical test performance (RMSE \textasciitilde 40 µg/m\textsuperscript{3}, $R^2 \approx$ 0.35), and all show a large train-to-test RMSE gap (\textasciitilde 25 train vs. \textasciitilde 40 test). The gap is larger than typical overfitting: the training set covers spring through autumn 2004, while the test set is winter 2004--2005. Seasonal changes in boundary-layer dynamics, photochemistry, and sensor behaviour mean the test distribution is systematically different from training. **OLS** is not badly overfit relative to the regularised alternatives, suggesting the eight predictors are informative but the main challenge is distributional shift, not in-sample variance. **Ridge** (CV-optimal $\lambda = `r sprintf("%.3f", ridge_lambda)`$) gives marginal improvement over OLS, confirming that coefficient instability from multicollinearity is not the binding constraint. **Lasso** (CV-optimal $\lambda = `r sprintf("%.4f", lasso_lambda)`$) retains `r length(retained)` of `r length(FEATURES)` predictors --- noteworthy because PC1 alone explains `r sprintf("%.0f%%", 100*var_exp[1])` of sensor variance, yet Lasso assigns nonzero weight to most sensors individually. Each retained sensor contributes marginal information about `NO2_GT` beyond what the others provide; `PT08_S3_NOx` (negative on PC1) is retained, confirming it adds predictive signal the variance-maximising PC1 direction partially suppresses. Additionally, that Lasso retains all predictors with nonzero weight indicates it is operating in a shrinkage rather than selection regime here, functioning more similarly to Ridge than is typical of $L_1$ regularization. **PC Regression** ($k = `r best_k`$) shows CV RMSE decreasing monotonically from $k = 1$ to $k = `r best_k`$, meaning every PC --- including PC3--PC5, which together explain only `r sprintf("%.0f%%", 100*(1 - cum_var[2]))` of sensor variance --- contributes useful information about `NO2_GT`. With $k = 5$, PC regression is algebraically equivalent to OLS on the scaled raw sensors (it spans the same column space), explaining identical test metrics. This is the core tension between PCA and Lasso: PCA discards low-variance PCs on structural grounds, but CV reveals those directions are predictively relevant. Given that CV-optimal performance requires all 5 PCs, PCR provides no dimensionality reduction benefit over OLS on this dataset and Lasso, despite retaining all predictors, is the preferred approach as it at least allows coefficient shrinkage guided by predictive relevance.

## Section 7: Limitations

1.  **Residual autocorrelation.** Hourly observations are strongly serially correlated. OLS standard errors assume independence, which they do not have. Newey--West (HAC) standard errors would correct inference without changing point estimates; this was not implemented. The CV folds also do not account for serial correlation, so reported CV RMSEs are optimistic.
2.  **Seasonal distributional shift / sensor drift.** The large train-test RMSE gap (\textasciitilde 25 vs. \textasciitilde 40) is not explained by overfitting alone --- it reflects systematic differences between training (spring--autumn 2004) and test (winter 2004--2005). Electrochemical sensors are known to drift over months of operation and respond differently under winter atmospheric chemistry.
3.  **Median imputation compresses tails.** The provided CSV replaced missing values with column medians. Imputing at the median artificially pulls extreme observations toward the centre, understating tail variance.
4.  **Single monitoring site.** All data come from one urban location in Italy over one year. Results may not generalise to other cities, sensor deployments, climates, or traffic profiles.
5.  **Reference measurement error.** GT variables carry calibration uncertainty and sampling noise. Treating `NO2_GT` as noiseless ground truth introduces errors-in-variables bias --- true predictive signal from the sensors may be somewhat higher than $R^2$ values suggest.
6.  **Log transform omitted.** Raw `NO2_GT` is mildly right-skewed (skewness 0.73), but the log transform over-corrects to a left-skewed distribution (skewness $-1.09$). A Box--Cox transformation could achieve better symmetry but was not explored.

\clearpage

# Code Appendix

All code used to produce the results above is reproduced here in the order it appears in the report. Chunks are not re-executed; this appendix is for inspection and reproducibility.

```{r appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
