---
title: "Predicting Urban NO\\textsubscript{2} Concentrations from Low-Cost Sensor Arrays"
subtitle: "STAT 24620 / FINM 34700 / STAT 32950 --- Group Data Project, Spring 2026"
author: "[Name 1], [Name 2], [Name 3]"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: false
    number_sections: false
    fig_caption: true
    keep_tex: false
geometry: margin=0.85in
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \floatplacement{figure}{H}
  - \setlength{\parskip}{2pt}
  - \setlength{\textfloatsep}{6pt plus 1pt minus 2pt}
  - \setlength{\intextsep}{6pt plus 1pt minus 2pt}
---

```{r setup, include=FALSE}
# Global knitr options: hide code throughout the body; show results, figures, tables.
# All code is reproduced in the Code Appendix at the end.
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 6.5,
  fig.height = 3,
  dpi       = 120,
  out.width = "92%"
)

# Packages
library(tidyverse)   # dplyr, ggplot2, tidyr, readr
library(lubridate)   # datetime arithmetic
library(patchwork)   # combining ggplots
library(glmnet)      # ridge + lasso
library(e1071)       # skewness()
library(zoo)         # rolling means
library(knitr)       # kable()

set.seed(42)
```

**AI acknowledgment:** Claude Code (Anthropic) was used for coding assistance and debugging. All analysis design, interpretation, and written explanations are the authors' own work.

## Research Question

Can a five-sensor electrochemical array (PT08\_S1--S5) accurately predict hourly NO\textsubscript{2} concentrations at an Italian urban monitoring site, and does dimensionality reduction via PCA improve prediction relative to directly regressing on the raw sensor readings? Air-quality monitoring increasingly relies on low-cost sensors alongside certified reference analyzers. If a small number of principal components captures most sensor variance, PC regression may be more stable than full sensor regression --- especially when sensors are collinear and subject to environmental drift. We compare OLS, Ridge, Lasso, and PC regression, all augmented with meteorological covariates, on a chronological hold-out test set.

## Dataset

**Source:** UCI Machine Learning Repository --- Air Quality Dataset (De Vito et al., 2008). Hourly measurements from March 2004 to April 2005 at one urban site in Italy. The project-provided CSV was partially pre-cleaned: heavily missing variables were dropped, rows without valid dates were removed, and remaining gaps were median-imputed. Residual $-200$ sentinel codes survive in `CO_GT` and `C6H6_GT`; these two variables are not used as predictors or as the response, so they do not affect any model results. **Variable groups:** *Sensors* `PT08_S1_CO`, `PT08_S2_NMHC`, `PT08_S3_NOx`, `PT08_S4_NO2`, `PT08_S5_O3`; *weather* `T` (°C), `RH` (%), `AH` (g/m\textsuperscript{3}); *reference* `CO_GT`, `C6H6_GT`, `NOx_GT`, `NO2_GT`; *response* `NO2_GT` (µg/m\textsuperscript{3}).

## Section 1: EDA and Preprocessing

```{r load-data, results='hide'}
df <- read.csv("C:/Users/MiahH/OneDrive/Documents/FINM-34700-HWs/air_quality_italy.csv",
               stringsAsFactors = FALSE)

# Diagnostic checks (suppressed in body; visible in Code Appendix)
cat("Shape:", nrow(df), "rows x", ncol(df), "cols\n")
print(sapply(df, class))
print(colSums(is.na(df)))

num_cols   <- names(df)[sapply(df, is.numeric)]
neg200_hit <- sapply(df[num_cols], function(x) any(x == -200, na.rm = TRUE))
neg200_in  <- names(neg200_hit)[neg200_hit]
if (length(neg200_in)) print(neg200_in) else cat("None -- clean.\n")

# Build proper datetime index
df$timestamp <- as.POSIXct(df$Date, tz = "UTC") + lubridate::hours(df$Hour)
df <- df[order(df$timestamp), ]
rownames(df) <- NULL

cat("\nDate range:", format(min(df$timestamp)), "to", format(max(df$timestamp)), "\n")
cat("Total hours:", nrow(df), "\n")
```

```{r fig1-eda, fig.width=7, fig.height=2.4, fig.cap="NO$_2$ reference concentration over time (left) and distribution on raw vs.\\ log scale (right)."}
df$NO2_roll <- zoo::rollmean(df$NO2_GT, k = 168, fill = NA, align = "right")

p_ts <- ggplot(df, aes(x = timestamp)) +
  geom_line(aes(y = NO2_GT, color = "Hourly"), linewidth = 0.2, alpha = 0.55) +
  geom_line(aes(y = NO2_roll, color = "7-day rolling mean"), linewidth = 0.7) +
  scale_color_manual(name = NULL,
                     values = c("Hourly" = "steelblue",
                                "7-day rolling mean" = "darkblue")) +
  labs(x = "Date", y = expression(NO[2]~GT~(mu*g/m^3)),
       title = expression(NO[2]~"Over Time")) +
  theme_minimal(base_size = 8) +
  theme(legend.position = "bottom",
        legend.margin = margin(0,0,0,0),
        legend.box.spacing = unit(2, "pt"),
        axis.text.x = element_text(angle = 30, hjust = 1),
        plot.title = element_text(size = 9))

log_vals <- log(df$NO2_GT[df$NO2_GT > 0 & !is.na(df$NO2_GT)])

p_dist <- ggplot() +
  geom_histogram(aes(x = df$NO2_GT, y = after_stat(density), fill = "Raw NO2_GT"),
                 bins = 60, alpha = 0.6) +
  geom_histogram(aes(x = log_vals * 30, y = after_stat(density) / 30,
                     fill = "log(NO2_GT) (rescaled)"),
                 bins = 60, alpha = 0.45) +
  scale_fill_manual(name = NULL,
                    values = c("Raw NO2_GT" = "steelblue",
                               "log(NO2_GT) (rescaled)" = "darkorange")) +
  labs(x = expression(NO[2]~GT~(mu*g/m^3)), y = "Density",
       title = expression("Distribution of"~NO[2]~GT)) +
  theme_minimal(base_size = 8) +
  theme(legend.position = "bottom",
        legend.margin = margin(0,0,0,0),
        legend.box.spacing = unit(2, "pt"),
        plot.title = element_text(size = 9))

p_ts + p_dist
```

The time-series panel shows strong diurnal cycling (morning and evening traffic peaks) and a seasonal drift: concentrations rise in autumn/winter when boundary-layer mixing is suppressed and fall in spring/summer. Raw `NO2_GT` is mildly right-skewed (`r sprintf("%.2f", e1071::skewness(df$NO2_GT, na.rm = TRUE))`), while the log transform over-corrects to a left-skewed distribution (`r sprintf("%.2f", e1071::skewness(log_vals, na.rm = TRUE))`). Since the log transform does not improve symmetry, we fit all supervised models on the raw concentration scale and note the choice as a limitation.

```{r fig2-corr, fig.width=6.5, fig.height=4.8, fig.cap="Pearson correlation heatmap across sensors (top-left block), reference analyzers (middle block), and weather (bottom-right block). Block separators are drawn in black."}
SENSORS <- c("PT08_S1_CO", "PT08_S2_NMHC", "PT08_S3_NOx", "PT08_S4_NO2", "PT08_S5_O3")
GT_VARS <- c("CO_GT", "C6H6_GT", "NOx_GT", "NO2_GT")
WEATHER <- c("T", "RH", "AH")
TARGET  <- "NO2_GT"

corr_cols <- c(SENSORS, GT_VARS, WEATHER)
corr_mat  <- cor(df[, corr_cols], use = "pairwise.complete.obs")

corr_df <- as.data.frame(as.table(corr_mat))
names(corr_df) <- c("Var1", "Var2", "value")
corr_df$Var1 <- factor(corr_df$Var1, levels = corr_cols)
corr_df$Var2 <- factor(corr_df$Var2, levels = rev(corr_cols))

block_breaks <- c(length(SENSORS), length(SENSORS) + length(GT_VARS))

ggplot(corr_df, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white", linewidth = 0.2) +
  geom_text(aes(label = sprintf("%.2f", value)), size = 2.1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  geom_vline(xintercept = block_breaks + 0.5, linewidth = 0.5) +
  geom_hline(yintercept = length(corr_cols) - block_breaks + 0.5, linewidth = 0.5) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        legend.key.width = unit(8, "pt"))
```

The block structure motivates both PCA (the sensor block is internally correlated, suggesting low intrinsic rank) and the supervised analysis (cross-block correlation between sensors and `NO2_GT` supports sensor-based prediction).

## Section 2: Chronological Train / Test Split

Because the data are hourly time-series with strong autocorrelation and seasonal drift, a random split would leak future information into the past. We use a chronological holdout: the first 75% form the training set, the remaining 25% form the test set. All preprocessing statistics (scaling means/SDs, PCA directions, regularization hyperparameters) are estimated on training data only and applied to test data without re-fitting.

```{r split}
n         <- nrow(df)
split_idx <- floor(0.75 * n)
train <- df[1:split_idx, ]
test  <- df[(split_idx + 1):n, ]

cat(sprintf("Training: %d rows (%s to %s).  Test: %d rows (%s to %s).  Split at %s.\n",
            nrow(train),
            format(min(train$timestamp), "%Y-%m-%d"),
            format(max(train$timestamp), "%Y-%m-%d"),
            nrow(test),
            format(min(test$timestamp), "%Y-%m-%d"),
            format(max(test$timestamp), "%Y-%m-%d"),
            format(min(test$timestamp), "%Y-%m-%d")))
```

## Section 3: PCA on the Sensor Block

We run PCA exclusively on the five PT08 sensor variables. Weather variables are excluded so that the structural story stays clean: the question here is *how many independent axes of variation exist in the sensor array itself?* Weather enters only as covariates in the supervised stage. Standardization (zero mean, unit variance) uses training-set statistics only and is applied identically to the test sensors.

```{r pca-fit, results='hide'}
train_sensors <- as.matrix(train[, SENSORS])
test_sensors  <- as.matrix(test[, SENSORS])

sens_mean <- colMeans(train_sensors)
sens_sd   <- apply(train_sensors, 2, sd)

train_sensors_sc <- scale(train_sensors, center = sens_mean, scale = sens_sd)
test_sensors_sc  <- scale(test_sensors,  center = sens_mean, scale = sens_sd)

pca <- prcomp(train_sensors_sc, center = FALSE, scale. = FALSE)
train_pcs <- pca$x
test_pcs  <- as.matrix(test_sensors_sc) %*% pca$rotation

var_exp <- pca$sdev^2 / sum(pca$sdev^2)
cum_var <- cumsum(var_exp)

loadings <- pca$rotation
colnames(loadings) <- paste0("PC", 1:ncol(loadings))

# Suppressed; figure and discussion convey the same information
cat("Variance explained per PC:\n")
for (i in seq_along(var_exp)) {
  cat(sprintf("  PC%d: %.3f  (cumulative: %.3f)\n", i, var_exp[i], cum_var[i]))
}
cat("\nLoadings matrix:\n")
print(round(loadings, 3))
```

```{r fig3-scree, fig.width=7, fig.height=2.6, fig.cap="Scree plot (left) and PC1/PC2 sensor loadings (right)."}
scree_df <- data.frame(PC = factor(paste0("PC", 1:5), levels = paste0("PC", 1:5)),
                       var = var_exp, cum = cum_var)

p_scree <- ggplot(scree_df, aes(x = PC)) +
  geom_col(aes(y = var), fill = "steelblue", alpha = 0.75) +
  geom_line(aes(y = cum, group = 1), color = "darkred", linewidth = 0.7) +
  geom_point(aes(y = cum), color = "darkred", size = 1.5) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "gray40") +
  annotate("text", x = 4.7, y = 0.92, label = "90%", color = "gray40", size = 2.5) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(y = "Variance explained", title = "Scree", x = NULL) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(size = 9))

sensor_labels <- c(
  "PT08_S1_CO"   = "S1\n(CO)",
  "PT08_S2_NMHC" = "S2\n(NMHC)",
  "PT08_S3_NOx"  = "S3\n(NOx)",
  "PT08_S4_NO2"  = "S4\n(NO2)",
  "PT08_S5_O3"   = "S5\n(O3)"
)

load_df <- as.data.frame(loadings[, 1:2])
load_df$sensor <- factor(rownames(load_df), levels = SENSORS)
load_long <- tidyr::pivot_longer(load_df, cols = c("PC1", "PC2"),
                                 names_to = "component", values_to = "loading")

p_load <- ggplot(load_long,
                 aes(x = sensor, y = loading, fill = component)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.85) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c("PC1" = "steelblue", "PC2" = "darkorange"),
                    name = NULL) +
  scale_x_discrete(labels = sensor_labels) +
  labs(y = "Loading", x = NULL, title = "PC1 / PC2 loadings") +
  theme_minimal(base_size = 8) +
  theme(legend.position = "bottom",
        legend.margin = margin(0,0,0,0),
        legend.box.spacing = unit(2, "pt"),
        plot.title = element_text(size = 9))

p_scree + p_load
```

**PC1 (`r sprintf("%.0f%%", 100*var_exp[1])` of variance)** loads positively and nearly equally on four of the five sensors ($\approx +0.42$ to $+0.47$), but *negatively* on `PT08_S3_NOx` ($-0.43$). This sign flip is physically meaningful: PT08\_S3 is a metal-oxide sensor tuned to NOx, which responds inversely to ambient NO\textsubscript{2} oxidising conditions compared to the other sensors. PC1 therefore represents the overall combustion/traffic pollution load, with PT08\_S3 providing a counteracting signal. **PC2 (`r sprintf("%.0f%%", 100*var_exp[2])` of variance)** is strongly dominated by `PT08_S4_NO2` ($+0.84$), with `PT08_S3_NOx` ($+0.39$) and `PT08_S5_O3` ($-0.38$) as secondary contributors --- reflecting the chemical antagonism between NO\textsubscript{2} and O\textsubscript{3} in urban photochemistry. Together PC1 and PC2 explain `r sprintf("%.0f%%", 100*cum_var[2])` of sensor variance; PC3--PC5 each capture less than 6%. The low-rank structure motivates PC regression: projecting onto the first few PCs substitutes correlated raw sensors with orthogonal summary axes, which can stabilise regression coefficients. Whether this leads to better out-of-sample prediction depends on whether the discarded PCs carry useful signal --- tested in Section 5.

## Section 4: Supervised Models

We fit four models predicting `NO2_GT` (raw scale), each using the five sensors plus `T`, `RH`, `AH` as predictors. All predictors are standardised using a separate scaler fit on training data only. **OLS** (baseline, no regularisation); **Ridge** ($L_2$ penalty, $\lambda$ by 5-fold CV); **Lasso** ($L_1$ penalty, $\lambda$ by 5-fold CV); **PC regression** (OLS on first $k$ PCs + weather, $k$ by 5-fold CV). This R implementation uses `glmnet`, whose penalty $\lambda$ is parameterised differently from `scikit-learn`'s $\alpha$ (glmnet scales the loss by $1/n$); optimal numerical values therefore differ from a Python implementation, but the methodology is identical.

```{r build-features, results='hide'}
FEATURES <- c(SENSORS, WEATHER)

X_train_raw <- as.matrix(train[, FEATURES])
X_test_raw  <- as.matrix(test[, FEATURES])

sup_mean <- colMeans(X_train_raw)
sup_sd   <- apply(X_train_raw, 2, sd)

X_train <- scale(X_train_raw, center = sup_mean, scale = sup_sd)
X_test  <- scale(X_test_raw,  center = sup_mean, scale = sup_sd)

y_train <- train[[TARGET]]
y_test  <- test[[TARGET]]
```

```{r fit-ols-ridge-lasso, results='hide'}
# --- OLS ---
ols_df_train <- data.frame(y = y_train, X_train)
ols_fit      <- lm(y ~ ., data = ols_df_train)
ols_train_pred <- predict(ols_fit, newdata = ols_df_train)
ols_test_pred  <- predict(ols_fit, newdata = data.frame(X_test))

# --- Ridge (5-fold CV on train) ---
ridge_cv <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 5,
                      type.measure = "mse", standardize = FALSE)
ridge_lambda     <- ridge_cv$lambda.min
ridge_train_pred <- as.numeric(predict(ridge_cv, newx = X_train, s = "lambda.min"))
ridge_test_pred  <- as.numeric(predict(ridge_cv, newx = X_test,  s = "lambda.min"))

# --- Lasso (5-fold CV on train) ---
lasso_cv <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 5,
                      type.measure = "mse", standardize = FALSE)
lasso_lambda     <- lasso_cv$lambda.min
lasso_train_pred <- as.numeric(predict(lasso_cv, newx = X_train, s = "lambda.min"))
lasso_test_pred  <- as.numeric(predict(lasso_cv, newx = X_test,  s = "lambda.min"))

lasso_coefs_full <- as.numeric(coef(lasso_cv, s = "lambda.min"))
names(lasso_coefs_full) <- c("(Intercept)", FEATURES)
lasso_coefs <- lasso_coefs_full[-1]

zeroed   <- names(lasso_coefs)[lasso_coefs == 0]
retained <- names(lasso_coefs)[lasso_coefs != 0]
```

```{r fit-pcr, results='hide'}
# --- PC Regression: choose k by 5-fold CV on train (preserving time order) ---
train_weather_raw <- as.matrix(train[, WEATHER])
test_weather_raw  <- as.matrix(test[, WEATHER])
w_mean <- colMeans(train_weather_raw)
w_sd   <- apply(train_weather_raw, 2, sd)
train_weather_sc <- scale(train_weather_raw, center = w_mean, scale = w_sd)
test_weather_sc  <- scale(test_weather_raw,  center = w_mean, scale = w_sd)

n_train  <- nrow(X_train)
fold_id  <- ceiling(seq_len(n_train) / (n_train / 5))
fold_id  <- pmin(fold_id, 5L)

cv_rmse <- numeric(5)
for (k in 1:5) {
  fold_rmse <- numeric(5)
  X_pc_full <- cbind(train_pcs[, 1:k, drop = FALSE], train_weather_sc)
  for (f in 1:5) {
    val_idx  <- which(fold_id == f)
    tr_idx   <- setdiff(seq_len(n_train), val_idx)
    fit      <- lm.fit(cbind(1, X_pc_full[tr_idx, , drop = FALSE]), y_train[tr_idx])
    pred     <- cbind(1, X_pc_full[val_idx, , drop = FALSE]) %*% fit$coefficients
    fold_rmse[f] <- sqrt(mean((y_train[val_idx] - pred)^2))
  }
  cv_rmse[k] <- mean(fold_rmse)
}

best_k <- which.min(cv_rmse)

X_pc_train_best <- cbind(train_pcs[, 1:best_k, drop = FALSE], train_weather_sc)
X_pc_test_best  <- cbind(test_pcs[,  1:best_k, drop = FALSE], test_weather_sc)
colnames(X_pc_train_best) <- c(paste0("PC", 1:best_k), WEATHER)
colnames(X_pc_test_best)  <- colnames(X_pc_train_best)

pcr_df_train <- data.frame(y = y_train, X_pc_train_best)
pcr_fit      <- lm(y ~ ., data = pcr_df_train)
pcr_train_pred <- predict(pcr_fit, newdata = pcr_df_train)
pcr_test_pred  <- predict(pcr_fit, newdata = data.frame(X_pc_test_best))
```

## Section 5: Comparison

```{r table1-results}
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

results <- data.frame(
  Model = c(
    "OLS (baseline)",
    sprintf("Ridge (lambda=%.3f)",  ridge_lambda),
    sprintf("Lasso (lambda=%.4f)",  lasso_lambda),
    sprintf("PC Regression (k=%d)", best_k)
  ),
  `Train RMSE` = c(rmse(y_train, ols_train_pred),
                   rmse(y_train, ridge_train_pred),
                   rmse(y_train, lasso_train_pred),
                   rmse(y_train, pcr_train_pred)),
  `Test RMSE`  = c(rmse(y_test, ols_test_pred),
                   rmse(y_test, ridge_test_pred),
                   rmse(y_test, lasso_test_pred),
                   rmse(y_test, pcr_test_pred)),
  `Test R2`    = c(r2(y_test, ols_test_pred),
                   r2(y_test, ridge_test_pred),
                   r2(y_test, lasso_test_pred),
                   r2(y_test, pcr_test_pred)),
  check.names = FALSE
)
results[, 2:4] <- round(results[, 2:4], 3)

kable(results,
      caption  = "Model comparison on the chronological hold-out test set.",
      booktabs = TRUE,
      col.names = c("Model", "Train RMSE", "Test RMSE", "Test $R^2$"),
      escape    = FALSE)
```

```{r fig4-pred-vs-actual, fig.width=6.5, fig.height=5, fig.cap="Predicted vs.\\ actual NO$_2$ on the chronological test set, by model."}
model_preds <- list(
  list(name = "OLS (baseline)",                              pred = ols_test_pred,   color = "steelblue"),
  list(name = sprintf("Ridge (lambda=%.3f)", ridge_lambda),  pred = ridge_test_pred, color = "forestgreen"),
  list(name = sprintf("Lasso (lambda=%.4f)", lasso_lambda),  pred = lasso_test_pred, color = "darkorange"),
  list(name = sprintf("PC Regression (k=%d)", best_k),       pred = pcr_test_pred,   color = "#DC143C")
)

plot_df <- do.call(rbind, lapply(model_preds, function(m) {
  data.frame(model     = m$name,
             color     = m$color,
             actual    = y_test,
             predicted = m$pred,
             stringsAsFactors = FALSE)
}))
plot_df$model <- factor(plot_df$model, levels = sapply(model_preds, function(m) m$name))

ann_df <- plot_df %>%
  group_by(model) %>%
  summarise(rmse_val = sqrt(mean((actual - predicted)^2)),
            r2_val   = 1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2),
            .groups  = "drop") %>%
  mutate(label = sprintf("RMSE = %.2f, R^2 = %.3f", rmse_val, r2_val))

lo <- min(c(plot_df$actual, plot_df$predicted)) - 5
hi <- max(c(plot_df$actual, plot_df$predicted)) + 5

ggplot(plot_df, aes(x = actual, y = predicted)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "black", linewidth = 0.3) +
  geom_point(aes(color = model), alpha = 0.18, size = 0.4, show.legend = FALSE) +
  scale_color_manual(values = setNames(sapply(model_preds, function(m) m$color),
                                       sapply(model_preds, function(m) m$name))) +
  facet_wrap(~ model, ncol = 2) +
  geom_text(data = ann_df,
            aes(x = lo + 5, y = hi - 10, label = label),
            hjust = 0, vjust = 1, size = 2.4, inherit.aes = FALSE) +
  coord_fixed(xlim = c(lo, hi), ylim = c(lo, hi)) +
  labs(x = expression("Actual"~NO[2]~GT),
       y = expression("Predicted"~NO[2]~GT)) +
  theme_minimal(base_size = 8) +
  theme(strip.text = element_text(face = "bold", size = 8),
        panel.spacing = unit(4, "pt"))
```

**Discussion.** All four models achieve nearly identical test performance (RMSE \textasciitilde 40 µg/m\textsuperscript{3}, $R^2 \approx$ 0.35), and all show a large train-to-test RMSE gap (\textasciitilde 25 train vs.\ \textasciitilde 40 test). The gap is larger than typical overfitting: the training set covers spring through autumn 2004, while the test set is winter 2004--2005. Seasonal changes in boundary-layer dynamics, photochemistry, and sensor behaviour mean the test distribution is systematically different from training. **OLS** is not badly overfit relative to the regularised alternatives, suggesting the eight predictors are informative but the main challenge is distributional shift, not in-sample variance. **Ridge** (CV-optimal $\lambda = `r sprintf("%.3f", ridge_lambda)`$) gives marginal improvement over OLS, confirming that coefficient instability from multicollinearity is not the binding constraint. **Lasso** (CV-optimal $\lambda = `r sprintf("%.4f", lasso_lambda)`$) retains `r length(retained)` of `r length(FEATURES)` predictors --- noteworthy because PC1 alone explains `r sprintf("%.0f%%", 100*var_exp[1])` of sensor variance, yet Lasso assigns nonzero weight to most sensors individually. Each retained sensor contributes marginal information about `NO2_GT` beyond what the others provide; `PT08_S3_NOx` (negative on PC1) is retained, confirming it adds predictive signal the variance-maximising PC1 direction partially suppresses. **PC Regression** ($k = `r best_k`$) shows CV RMSE decreasing monotonically from $k = 1$ to $k = `r best_k`$, meaning every PC --- including PC3--PC5, which together explain only `r sprintf("%.0f%%", 100*(1 - cum_var[2]))` of sensor variance --- contributes useful information about `NO2_GT`. With $k = 5$, PC regression is algebraically equivalent to OLS on the scaled raw sensors (it spans the same column space), explaining identical test metrics. This is the core tension between PCA and Lasso: PCA discards low-variance PCs on structural grounds, but CV reveals those directions are predictively relevant.

## Section 6: Limitations

1. **Residual autocorrelation.** Hourly observations are strongly serially correlated. OLS standard errors assume independence, which they do not have. Newey--West (HAC) standard errors would correct inference without changing point estimates; this was not implemented. The CV folds also do not account for serial correlation, so reported CV RMSEs are optimistic.
2. **Seasonal distributional shift / sensor drift.** The large train-test RMSE gap (\textasciitilde 25 vs.\ \textasciitilde 40) is not explained by overfitting alone --- it reflects systematic differences between training (spring--autumn 2004) and test (winter 2004--2005). Electrochemical sensors are known to drift over months of operation and respond differently under winter atmospheric chemistry.
3. **Median imputation compresses tails.** The provided CSV replaced missing values with column medians. Imputing at the median artificially pulls extreme observations toward the centre, understating tail variance.
4. **Single monitoring site.** All data come from one urban location in Italy over one year. Results may not generalise to other cities, sensor deployments, climates, or traffic profiles.
5. **Reference measurement error.** GT variables carry calibration uncertainty and sampling noise. Treating `NO2_GT` as noiseless ground truth introduces errors-in-variables bias --- true predictive signal from the sensors may be somewhat higher than $R^2$ values suggest.
6. **Log transform omitted.** Raw `NO2_GT` is mildly right-skewed (skewness 0.73), but the log transform over-corrects to a left-skewed distribution (skewness $-1.09$). A Box--Cox transformation could achieve better symmetry but was not explored.

\clearpage

# Code Appendix

All code used to produce the results above is reproduced here in the order it appears in the report. Chunks are not re-executed; this appendix is for inspection and reproducibility.

```{r appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
